---
title: 'Scripts for Making the GUSG AIM worksheet for 2023'
author: 'Michael Schmidt'
date: '`r Sys.Date()`'
output:
  pagedown::html_paged:
    self_contained: false
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning=FALSE
)
```

```{r}
options(
  max.print=150,
  vsc.use_httpgd = TRUE,
  device = 'windows'
)
library(tidyverse)
library(sf)
library(trfotools)
library(here)
library(terra)
library(tidyselect)
library(MetBrewer)

theme_set(theme_minimal())
```

## Load the AIM and ESG data
```{r}
terradat <- readRDS(here("data/cache/terradat_2018to2021.rds"))
esg<-rast("G:\\r_scripts\\aim\\projects\\ecological_site_groups\\data\\ESGs\\ESGs.tif")

# Turn on ESG layer instead of count
activeCat(esg)<-"ESG"
```

## Clip to TRFO and extract ESG to trfo points and reform for comparision
```{r}
tdat_trfo_cl<-terradat$tdat%>%
  st_transform(st_crs(trfo_b))%>%
  st_intersection(trfo_b)%>%
  st_transform(crs(esg))
  
esg_tdat_trfo<-terra::extract(esg, tdat_trfo_cl)%>%
  select(ESG)%>%
  cbind(tdat_trfo_cl)%>%
  as_tibble()%>%
  select(-geometry)

rm(terradat)
rm(esg)
```

## Load benchmarks
```{r}
bm_long<-read_csv("projects/2023_design_worksheet/output/long_benchmarks_2023.csv")

indicators<-bm_long%>%
  distinct(indicator_parent)%>%
  pull(indicator_parent)
```

## Transform ter
```{r}
long_esg_tdat_trfo<-esg_tdat_trfo%>%
  mutate(
    pct_sagebrush_dead=((AH_SagebrushCover-AH_SagebrushCover_Live)/AH_SagebrushCover)*100,
    FH_WoodyAndHerbLitterCover = FH_WoodyLitterCover+FH_HerbLitterCover
  )%>%
  select(ESG, PlotID, all_of(indicators))%>%
  pivot_longer(cols=all_of(indicators), names_to="indicator", values_to="point_value")

bm_long_cl<-bm_long%>%
  filter(!is.na(directional_indicator), directional_indicator != "n")%>%
  select(ESG, indicator_parent, indicator_cat=indicator, directional_indicator, bm_value=value)

sub_bm_comp_pts_trfo<-long_esg_tdat_trfo%>%
  left_join(bm_long_cl, by=c("ESG", "indicator"="indicator_parent"))

comparisons<-sub_bm_comp_pts_trfo%>%
  select(ESG, indicator, directional_indicator, bm_value)%>%
  group_by(ESG, indicator, directional_indicator)%>%
  summarize(bm_value = max(bm_value))%>%
  ungroup()%>%
  pivot_wider(names_from=directional_indicator, values_from=bm_value)

sub_bm_comp_pts_trfo%>%
  select(ESG, PlotID, indicator, point_value)%>%
  group_by()


#   mutate(indicator_comp = case_when(
#     directional_indicator == ">=" && point_value >= bm_value
#   ))%>%
#   count(ESG, PlotID, indicator, indicator_comp)%>%
#   filter(!(indicator_comp=="meets" & n==1))%>%
#   count(indicator, indicator_comp)


# bm_comp_pts_trfo%>%
#   ggplot(aes(indicator, n, fill=indicator_comp))+
#   geom_col()+
#   coord_flip()+
#   scale_fill_manual(values = c("#E74C3C", "#3498DB", "#2ECC71"))


var_low_marginal=c(
  "BareSoilCover", 
  "GapCover_25_plus",
  "GapCover_101_200",
  "GapCover_200_plus",
  "AH_SagebrushCover"
)
var_high_marginal=c(
  "TotalFoliarCover",
  "AH_NonNoxPerenForbCover",
  "Hgt_PerenForb_Avg",
  "AH_NonNoxPerenGrassCover",
  "Hgt_NonNoxPerenGrass_Avg",
  "Hgt_Sagebrush_Avg",
  "AH_NonNoxTreeCover",
  "FH_WoodyAndHerbLitterCover"
)

var_low_suitable=c(
  "AH_NoxCover",
  "AH_NoxAnnGrassCover",
  "pct_sagebrush_dead",
  "NumSpp_NoxPlant"
)

var_high_suitable=c(
  "SoilStability_Protected",
  "SoilStability_Unprotected"
)

conditional_bm_comp_trfo<-bm_comp_pts_trfo%>%
  mutate(conditional_rating = case_when(
    indicator %in% var_low_marginal & indicator_comp == "lower" ~ "Marginal",
    indicator %in% var_low_marginal & indicator_comp == "meets" ~ "Suitable",
    indicator %in% var_low_marginal & indicator_comp == "higher" ~ "Unsuitable",
    indicator %in% var_high_marginal & indicator_comp == "lower" ~ "Unsuitable",
    indicator %in% var_high_marginal & indicator_comp == "meets" ~ "Suitable",
    indicator %in% var_high_marginal & indicator_comp == "higher" ~ "Marginal",
    indicator %in% var_low_suitable & indicator_comp == "lower" ~ "Suitable",
    indicator %in% var_low_suitable & indicator_comp == "meets" ~ "Marginal",
    indicator %in% var_low_suitable & indicator_comp == "higher" ~ "Unsuitable",
    indicator %in% var_high_suitable & indicator_comp == "lower" ~ "Unsuitable",
    indicator %in% var_high_suitable & indicator_comp == "meets" ~ "Suitable",
    indicator %in% var_high_suitable & indicator_comp == "higher" ~ "Marginal",
  ))
```

```{r}
conditional_bm_comp_trfo%>%
  filter(!is.na(conditional_rating))%>%
  group_by(indicator)%>%
  mutate(
    freq=round(n/sum(n)*100, digits=1),
    conditional_rating=fct_relevel(conditional_rating, c("Unsuitable", "Marginal", "Suitable"))
  )%>%
  ggplot(aes(indicator, freq, fill=conditional_rating))+
  geom_col()+
  coord_flip()+
  scale_fill_manual(values = c("#E74C3C", "#3498DB", "#2ECC71" ))
```

I think if I had to do this over again I would keep the averages in here so that we could separate things better.  For example "NumSpp_NoxPlant" Suitable should be lowest 50% quantile marginal should be 50% to 75% and Unsuitable should be > 75%.

Also I looked at the difference of 50% quantile and mean and they are different.  This could be a future improvement. 
